\vspec{} and Friends: A suite of utilities to model spectroscopic phase curves of 3D exoplanet atmospheres in the presence of stellar variability.

### 2024-07-22 src/tex/ms.tex
Changes from Avi/Ted editing on July 22 2024

The \vspec{} stellar model is designed modularly to allow for both simple and complex behaviors. The model is available as the Python package \texttt{vspec-vsm} \TJ{(VSPEC Variable Star Model)}. It is spectral-model agnostic, meaning that it relies on another package (GridPolator, see Section \ref{sec:gridpolator}) to compute the actual variable spectra. Instead, the surface model takes a pair of latitude and longitude points as input and returns a dictionary $\{ T_{\rm eff} : f_{T_{\rm eff}} \}$ where the keys \teff~ are the effective temperatures of each surface component and the values $f_{T_{\rm eff}}$ are the fraction of the visible stellar disk with that effective temperature.

The main model class is the \texttt{Star} object, which describes bulk properties like radius, nominal \teff, and limb darkening parameters, in addition to acting as 
a container for objects representing sources of variability.

\subsection{Stellar Surface Grid}
Sources of variability that exist as major structures on the stellar surface (i.e. spots and faculae) are encoded as deviations from the nominal effective temperature on a grid of points representing the stellar surface. Each time the fractions of each surface component are calculated, the grid is initialized with constant \teff. The code then iterates through each spot and facula, replacing portions of the grid with the effective temperatures that make up each structure.

The simplest grid to build is one that is rectangular in latitude and longitude, and in which the area that each point represents decreases towards the poles. However, this can lead to unnecessary spatial oversampling at the poles, and may lead to poor resolution at the equator to keep the total number of points from being excessively high. To mitigate these problems \texttt{vspec-vsm} supports a ``grid'' generated using a Fibbonacci spiral to create a near-isotropic distribution of points. This is especially useful when simulating transits, giving optimal sampling at the stellar equator. It is also useful for resolving small structures on the surface, as the number of points needed for a given minimum resolution are greatly reduced.

\begin{figure}
    \centering
        \includegraphics[width=0.5\textwidth]{figures/surface_map_and_lc.pdf}
    \script{surface_map_and_lc.py}
    \caption{{\bf Top}: The surface map of a star with spots during a transit.
    The large elipses are spots, while the transiting planet occults a portion of the disk represented by the black circle.
    {\bf Bottom}: An example lightcurve of a spotted star.}
    \label{fig:surface_map}
\end{figure}

\subsection{Spots \label{subsec:spots}}
Our star spot model is nearly entirely based on observations of the sun. Sun spots can be resolved and are well-studied,
whereas spots on other stars (e.g., M dwarfs) can only be observed indirectly. We therefore designed our spot model to mimic
sun spots but with parameterized values for spot temperature and lifetime that can be matched to observations of other stellar types and ages.

On the Sun, spots have two regions shown if Figure \ref{fig:surface_map}: the dark, central umbra and the lighter, surrounding penumbra. A detailed review of sunspot behavior,
including sizes and lifetimes, is \citet{solanki2003}, however the specific solar values are not relevant to this description. However,
the spot area as a function of time is
\begin{equation}
    A(t) = \left\{
    \begin{array}{lr}
        A_0 e^{(t-t_0)/\tau}, & \text{if } t \leq t_0 \\
        A_0 - W(t-t_0), & \text{if } t > t_0
    \end{array}
    \right\}
\end{equation}
where $A_0$ is the maximum area reached, $t_0$ is the time of the maximum, $\tau$ is the exponential growth rate, and $W$ is the linear decay rate.

In addition to modeling the variability produced by a population of spots, \texttt{vspec-vsm} contains utilities to produce those populations given some description.
These parameters include the average spot area \citep[lognormally distributed, ][]{bogdan1988}, \teff, growth and decay rates, and the method for distributing spots on the surface. These \texttt{SpotGenerator} objects are responsible for evolving the spot population as the star ages. The number of spots created in some interval $\Delta t$ is:
\begin{equation}
    N(\Delta t) = \frac{4 \pi R_*^2 f_{\rm spot}}{A_{\rm mean}} \frac{\Delta t}{T}
\end{equation}
where $R_*$ is the stellar radius, $f_{\rm spot}$ is the fraction of the surface covered by spots in growth-decay equilibrium, $A_{\rm mean}$ is the lifetime-averaged mean spot area, and $T$ is the spot lifetime, defined as:
\begin{equation}
    T = \frac{A_0}{W} - \frac{1}{\tau} \ln{\left (\frac{A_{\rm min}}{A_0}\right )}
\end{equation}
where $A_{\rm min}$ is the area given to each spot when it is initialized.

There are also two modes to distribute spots on the star's surface. The first distributes them isotropically. The second, modeled after analysis of spots on the Sun by \citet{mandal2017}, concentrates the spot latitude around $\pm 15^{\circ}$.

\subsection{Faculae \label{subsec:faculae}}
Faculae are magnetically-generated regions of the solar surface that usually appear as bright points near the limb; we employ the ``hot wall''
model \citep{spruit1976} where faculae are described as three-dimensional pores in the stellar surface with a hot, bright wall, and a
cool, dark floor, as shown in Figure \ref{fig:fac_struct}.

Their three-dimensional structure causes faculae's observational properties to change depending on their angle from disk-center. Close to the limb,
the hot wall is visible to the observer, and faculae appear as bright points. Near the center, however, the cool floor is exposed and
faculae appear dark. To consider this effect in the faculae lightcurve, we compute the fraction of the facula's normalized area -- the
area on the disk it would occupy as a flat spot -- that is occupied by each the hot wall and cool floor. This is done via numerical integral
along the radius of the spot.

\begin{equation}
    f_{wall} = \frac{\int _{-R}^{R} Z_{\rm eff} dr}{\int _{-R}^{R} Z_{\rm eff} dr + \int _{-R}^{R} R_{\rm eff} dr}
\end{equation}
where $Z_{\rm eff}$ is the apparent height of the wall
\begin{equation}
    Z_{\rm eff} = \left\{
    \begin{array}{lr}
        Z_w \sin{\alpha}, & \text{if } \alpha \leq \alpha_{\rm crit} \\
        2\sqrt{R^2 - r^2}\cos{\alpha}, & \text{if } \alpha > \alpha_{\rm crit}
    \end{array}
    \right\}
\end{equation}
and $R_{\rm eff}$ is the apparent width of the floor
\begin{equation}
    R_{\rm eff} = \left\{
    \begin{array}{lr}
        2\sqrt{R^2 - r^2} - Z_w\sin{\alpha}, & \text{if } \alpha \leq \alpha_{\rm crit} \\
        0, & \text{if } \alpha > \alpha_{\rm crit}
    \end{array}
    \right\}
\end{equation}
for facula radius $R$, depth $Z_w$, and angle from disk-center $\alpha$. $r$ in this numerical scheme
is defined as the distance from the center of the facula along the radial line connecting the facula center to the
disk center. $\alpha_{\rm crit}$ is the value of alpha at which the floor is no longer visible and is defined to be $\arctan{\frac{2\sqrt{R^2-r^2}}{Z_w}}$.

According to \citet{topka1997}, faculae temperatures (of both the floor and wall) are dependent on the facula radius,
while depth appears to be constant. They also find that the smallest faculae have no visible floor, and that even at disk center they
appear as bright points. We parameterize the floor temperature to be
\begin{equation}
    \Delta T_{\rm eff, floor} = \left\{
    \begin{array}{lr}
    \text{Not visible}, & \text{if } r<r_{\rm min} \\
    m_{\rm floor}(r-r_{\rm min}) + \Delta T_{\rm eff, floor, 0}, & \text{if } r \ge r_{\rm min}
    \end{array}
    \right\}
\end{equation}

Where $r$ is the radius of the facula, $r_{\rm min}$ is the minimum radius where the floor is visible, $\Delta  T_{\rm eff, floor, 0}$ is
the difference between the floor and photosphere \teff at $r_{\rm min}$, and $m_{\rm floor}$ is the slope of the relationship with
units of [temperature] [length]$^{-1}$.

Similarly, the wall temperature is parameterized

\begin{equation}
    \Delta T_{\rm eff,wall} = m_{\rm wall}r + \Delta T_{\rm eff,wall,0}
\end{equation}

Where $\Delta T_{\rm eff,wall,0}$ is the temperature of a zero-radius facula and and $m_{\rm wall}$ is the slope of the radius-temperature
relationship with units of [temperature] [length]$^{-1}$.
These relationships can be defined by the user, for example setting $m=0$ for constant temperatures.

Facula lifetimes are defined as the time it takes to decay by $e^{-2}$. Because faculae grow and decay exponentially
at the same rate, each facula spends 1 lifetime with a radius greater than $e^{-1}$ of it's maximum. Each facula is born
and dies at $e^{-2}$ of its maximum, effectively existing in the code for two lifetimes.

\citet{hovis-afflerbach2022} suggest a typical facula lifetime to be on the order of 6 hours,
with a distribution that resembles a Poisson function. However, we choose a lognormal distribution
for both lifetime and maximum radius because it does not allow for these values to be 0. We choose to
correlate lifetime and radius so that they are determined by the normally distributed random variable $\mu$;
for each new facula, $\mu$ is randomly drawn and determines both the faclue lifetime and maximum radius.

\begin{figure}
    \centering
    \gridline{\includegraphics[width=0.5\textwidth]{figures/facula_model.png}}
    \gridline{\includegraphics[width=0.5\textwidth]{figures/facula_depth.pdf}}
    \script{facula_depth.py}
    \caption{
        ``Hot Wall'' model of faculae. Faculae structure causes their contrast to be dependent on their distance
        from the center of the disk. {\bf Top}: Depiction of a facula on the limb of a star. The hot wall is exposed
        to the observer causing the pore to appear bright. At disk center, the cool floor is most visible. {\bf Bottom}:
        The effects of depression depth and viewing angle on facula brightness. The 3D structure of faculae is most apparent
        when radius $\sim$ depth. A toy flux model was used to demonstrate the shape of these curves, but their magnitudes
        in practice depend on stellar spectral models.
        }
    \label{fig:fac_struct}
\end{figure}

\subsection{Flares \label{subsec:flares}}
Flares are an important source of stellar variability on short timescales. \texttt{vspec-vsm}'s flare lightcurve model is based on the \texttt{xoflares} package \citep{barclay2020}, which itself is based on empirically derived lightcurve shapes from \citet{davenport2016}. However, \texttt{xoflares} is designed to fit lightcurves in a single spectral band, so we add additional parameters to extend it to a multiwavelength lightcurve. We completely describe a flare using its temperature, total energy, full width at half maximum (FWHM), and the time of its peak. We model a flare as a hot, optically thin region above the photosphere that produces a blackbody spectrum. In this model, the temperature is constant, and the sharp rise and fall see in the lightcurve is caused by the region's rapidly changing area. This simplification allows the total flare energy $E$ to be used as a normalization factor:
\begin{equation}
    \int_{-\infty}^{\infty}A\,dt = \frac{E}{\sigma T^4}
\end{equation}
where $A$ is the time-dependent area of the flare region, $T$ is the constant flare temperature, and $\sigma$ is the Stefan-Boltzmann constant. This allows flare lightcurves to be produced with a fixed bolometric luminosity.

Because the temperature is fixed, the relevant quantity for modeling the flux of a flare in a given integration is the integrated time-area $\int A\,dt$. These quantities are computed numerically and fed to \vspec{} along with $T$ to produce spectra with the appropriate absolute flux.

Before simulating an observation, \vspec{} asks \texttt{vspec-vsm} to pre-compute a population of flares based on the frequency-energy relationship from \citet{gao2022}:
\begin{equation} \label{eq:flare_freq}
    \log{(f/{\rm [day]})} = \beta + \alpha~\log{(E/{\rm [erg]})}
\end{equation}

Where $f$ is the frequency of flares with energies $\ge E$.
We compute the number of expected flares over a time duration and determine the number to create $N$ by a random Poisson draw.
We then generate $N$ flares with energies determined by the quantile function:
\begin{equation}
    E = E_{\text{min}} (1-X)^{1/\alpha}
\end{equation}

where $E_{\text{min}}$ is the minimum considered flare energy and $X$ is a random number on the interval $[0,1)$.
We set the default values $\alpha=-0.829$, $\beta=26.87$ from \citet{gao2022}. Figure \ref{fig:flare_freq}
shows our simulation results compared to their powerlaw.

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figures/flare_freq.pdf}
    \caption{
        {\bf Top}: Generated flare frequencies compared to those expected from \citet{gao2022}, generated using a 10,000 day simulation.
        The $y$-axis shows the frequency of flares with energies {\em greater than or equal to} the value of the $x$-axis. Error bars
        based on the square root of the number of observed flares as this is a Poisson process.
        {\bf Bottom}: Lightcurve of a star flaring with the same powerlaw slope as the top panel, but the intercept ($\beta$) has been increased by 0.3
        for visual effect. The mean flare temperature is $9000$ K and the mean FWHM is $3$ hours.
        }
    \script{flare_freq.py}
    \label{fig:flare_freq}
\end{figure}

\subsection{Granulation}
Granulation is a source of stellar variability that arises from convection near the surface of the star.
The result is a stellar surface that is not constant in temperature and that changes on very short timescales.
Hydrodynamic simulations of stellar atmospheres \citep[e.g.][]{magic2014} describe hot granules of rising gas surrounded by cooler,
sinking regions with a \teff a few percent lower than the nominal value.

We model granulation as a global process, and its effects are computed after the effects of spots and faculae.
Of the remaining ``quiet'' photosphere (i.e. the regions not covered in spots of faculae), a fraction is computed to be part of the cool
inter-granule surface. The surface coverage of the cool region at any given time is computed by a Gaussian process (GP) using
the {\sc tinygp} package \citep{foreman-mackey2024} following the methodology of \citet{gordon2020}.
The GP uses a custom kernel function based on a power spectrum \citep{anderson1990,kallinger2014} to produce random changes in the coverage.

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figures/granulation_lightcurve.pdf}
    \caption{
        Effect of granulation on the lightcurve of a star. This simulation assumes a mean
        10\% coverage by inter-granule region whose temperature is 300 K lower than the surrounding photosphere.
        It also assumes that the coverage value varies with a magnitude of 1\% on a timescale of 6 hours.
        }
    \script{granule_lc.py}
    \label{fig:gran_lc}
\end{figure}

\subsection{Transits}
In the case of a transiting geometry, \texttt{vspec-vsm} is responsible for computing the properties of the portion of the stellar disk that is occulted. It does this by first projecting both the transiting planet and the visible part of the surface orthographically as they would be viewed by a distant observer. It then selects all the points that are near enough to the transit to be occulted. The code then iterates through each of those points and computes the fraction of it covered by the occulting circle using a 2D numerical integral.

@@

\section{Simple climate models}
\vspec{} allows the user to upload any desired GCM. However, for questions that focus on detectability,
the specific physics of the GCM are less important than properties like day/night temperature, surface pressure, or
the major absorbers in the atmosphere. To simplify these studies, \vspec{} has a built-in GCM that takes in just a
few parameters and produces a simple-yet-robust exoplanet climate. Simplicity allows the user to know everything about
the model and reduce surprises, but its basis in thermodynamics and robust energy balance make this model fit for publication-level science \TJ{is this really true? probably not}.

>>>>

The \vspec{} stellar model is designed in a modular fashion to allow for both simple and complex behaviors. The model is available as the Python package \texttt{vspec-vsm}. It is spectral-model agnostic, meaning that it relies on another package (GridPolator, see Section \ref{sec:gridpolator}) to compute the actual variable spectra. Instead, the surface model takes a pair of latitude and longitude points as input and returns a dictionary $\{ T_{\rm eff} : f_{T_{\rm eff}} \}$ where the keys \teff~ are the effective temperatures of each surface component and the values $f_{T_{\rm eff}}$ are the fraction of the visible stellar disk with that effective temperature.

The main model class is the \texttt{Star} object, which describes bulk properties (e.g. radius, photospheric \teff, limb darkening parameters) in addition to acting as 
a container for objects representing sources of variability. \TJ{Additional classes are defined for additional stellar components, e.g. spots, faculae,..., which are discussed below.}

\subsection{Stellar Surface Grid}
Sources of variability that exist as major structures on the stellar surface (i.e. spots and faculae) are encoded as deviations from the photospheric effective temperature on a grid of points representing the stellar surface. Each time the fractions of each surface component are calculated, the grid is initialized with constant \teff. The code then iterates through each spot and facula, replacing portions of the grid with the effective temperatures that make up each structure.

The simplest grid to build is one that is rectangular in latitude and longitude, and in which the area that each point represents decreases towards the poles. However, this can lead to unnecessary spatial oversampling at the poles, and may lead to poor resolution at the equator to keep the total number of points from being excessively high. To mitigate these problems \texttt{vspec-vsm} supports a ``grid'' generated using a Fibbonacci spiral to create a near-isotropic distribution of points. This is especially useful when simulating transits, giving optimal sampling at the stellar equator. It is also useful for resolving small structures on the surface, as the number of points needed for a given minimum resolution is greatly reduced.

\begin{figure}
    \centering
        \includegraphics[width=0.5\textwidth]{figures/surface_map_and_lc.pdf}
    \script{surface_map_and_lc.py}
    \caption{{\bf Top}: The surface map of a star with spots during a transit.
    The large ellipses are spots, while the transiting planet occults a portion of the disk represented by the black circle.
    {\bf Bottom}: An example lightcurve of a spotted star.
    \TJ{Remove transit and put it in a LD figure}}
    \label{fig:surface_map}
\end{figure}

\subsection{Spots \label{subsec:spots}}
Our star spot model is nearly entirely based on observations of the Sun. Sunspots can be resolved and are well-studied, whereas spots on other stars (especially non-solar-type stars) can only be observed indirectly. We therefore designed our spot model to mimic the behavior of sunspots but with parameterized values for spot temperature and lifetime that can be matched to observations of other stellar types and ages. On the Sun, spots have two regions shown in Figure \ref{fig:surface_map}: the dark, central umbra and the lighter, surrounding penumbra (a detailed review of sunspot behavior, including sizes and lifetimes, is described in \citet{solanki2003}). The spot area as a function of time is
\begin{equation}
    A(t) = \left\{
    \begin{array}{lr}
        A_0 e^{(t-t_0)/\tau}, & \text{if } t \leq t_0 \\
        A_0 - W(t-t_0), & \text{if } t > t_0
    \end{array}
    \right\}
\end{equation}
where $A_0$ is the maximum area reached, $t_0$ is the time of the maximum, $\tau$ is the exponential growth rate, and $W$ is the linear decay rate.

In addition to modeling the variability produced by a population of spots, \texttt{vspec-vsm} contains utilities to produce those populations given some description.
These parameters include the average spot area \citep[lognormally distributed, ][]{bogdan1988}, \teff, growth and decay rates, and the method for distributing spots on the surface. These \texttt{SpotGenerator} objects are responsible for evolving the spot population as the star ages. The number of spots created in some interval $\Delta t$ is:
\begin{equation}
    N(\Delta t) = \frac{4 \pi R_*^2 f_{\rm spot}}{A_{\rm mean}} \frac{\Delta t}{T}
\end{equation}
where $R_*$ is the stellar radius, $f_{\rm spot}$ is the fraction of the surface covered by spots in growth-decay equilibrium, $A_{\rm mean}$ is the lifetime-averaged mean spot area, and $T$ is the spot lifetime, defined as:
\begin{equation}
    T = \frac{A_0}{W} - \frac{1}{\tau} \ln{\left (\frac{A_{\rm min}}{A_0}\right )}
\end{equation}
where $A_{\rm min}$ is the area given to each spot when it is initialized. There are also two modes to distribute spots on the star's surface. The first distributes them isotropically; the second, modeled after analysis of spots on the Sun by \citet{mandal2017}, concentrates the spot latitude around $\pm 15^{\circ}$.

\subsection{Faculae \label{subsec:faculae}}
Faculae are magnetically-generated regions of the solar surface that usually appear as bright points near the limb; we employ the ``hot wall''
model \citep{spruit1976} where faculae are described as three-dimensional pores in the stellar surface with a hot, bright wall and a cool, dark floor, as shown in Figure \ref{fig:fac_struct}. Their three-dimensional structure causes faculaes' observational properties to change depending on their angle from disk-center. Close to the limb,
the hot wall is visible to the observer, and faculae appear as bright points; near the center, however, the cool floor is exposed and faculae appear dark. To consider this effect in the faculae lightcurve, we compute the fraction of each facula's projected area -- the area on the disk it would occupy as seen by the observer -- that is occupied by the hot wall versus the cool floor. This is done via numerical integral along the radius of the spot.

We calculate each portion using geometrical relationships between the spot dimensions and the viewing angle:
\begin{equation}
    f_{wall} = \frac{\int _{-R}^{R} Z_{\rm eff} dr}{\int _{-R}^{R} Z_{\rm eff} dr + \int _{-R}^{R} R_{\rm eff} dr}
\end{equation}
where $Z_{\rm eff}$ is the apparent height of the wall
\begin{equation}
    Z_{\rm eff} = \left\{
    \begin{array}{lr}
        Z_w \sin{\alpha}, & \text{if } \alpha \leq \alpha_{\rm crit} \\
        2\sqrt{R^2 - r^2}\cos{\alpha}, & \text{if } \alpha > \alpha_{\rm crit}
    \end{array}
    \right\}
\end{equation}
and $R_{\rm eff}$ is the apparent width of the floor
\begin{equation}
    R_{\rm eff} = \left\{
    \begin{array}{lr}
        2\sqrt{R^2 - r^2} - Z_w\sin{\alpha}, & \text{if } \alpha \leq \alpha_{\rm crit} \\
        0, & \text{if } \alpha > \alpha_{\rm crit}
    \end{array}
    \right\}
\end{equation}
for facula radius $R$, depth $Z_w$, and angle from disk-center $\alpha$. $r$ in this numerical scheme is defined as the distance from the center of the facula along the radial line connecting the facula center to the disk center. $\alpha_{\rm crit}$ is the value of alpha at which the floor is no longer visible and is defined to be $\arctan{\frac{2\sqrt{R^2-r^2}}{Z_w}}$.

According to studies of solar faculae \citep{topka1997}, faculae temperatures (of both the floor and wall) are dependent on the facula radius, while depth appears to be constant. They also find that the smallest faculae have no visible floor, and that even at disk center they appear as bright points. We parameterize the floor temperature to be
\begin{equation}
    \Delta T_{\rm eff, floor} = \left\{
    \begin{array}{lr}
    \text{Not visible}, & \text{if } r<r_{\rm min} \\
    m_{\rm floor}(r-r_{\rm min}) + \Delta T_{\rm eff, floor, 0}, & \text{if } r \ge r_{\rm min}
    \end{array}
    \right\}
\end{equation}
\TJ{Change variable names, make diagram}
where $r$ is the radius of the facula, $r_{\rm min}$ is the minimum radius where the floor is visible, $\Delta  T_{\rm eff, floor, 0}$ is the difference between the floor and photosphere \teff at $r_{\rm min}$, and $m_{\rm floor}$ is the slope of the relationship with units of [temperature] [length]$^{-1}$. Similarly, the wall temperature is parameterized as
\begin{equation}
    \Delta T_{\rm eff,wall} = m_{\rm wall}r + \Delta T_{\rm eff,wall,0}
\end{equation}
where $\Delta T_{\rm eff,wall,0}$ is the temperature of a zero-radius facula and $m_{\rm wall}$ is the slope of the radius-temperature relationship with units of [temperature] [length]$^{-1}$. These relationships can be defined by the user, for example setting $m=0$ for constant temperatures.

Facula lifetimes are defined as the time it takes its radius to decay by $e^{-2}$. Because faculae grow and decay exponentially at the same rate, each facula spends one lifetime with a radius greater than $e^{-1}$ of its maximum. Each facula is born and dies at a radius of $e^{-2}$ of its maximum, effectively existing in the code for two lifetimes. \citet{hovis-afflerbach2022} suggest a typical facula lifetime to be on the order of 6 hours, with a distribution that resembles a Poisson function; however, we choose a lognormal distribution for both lifetime and maximum radius because it does not allow for these values to be 0. We choose to correlate lifetime and radius so that they are determined by a normalized random distribution - for each new facula, a value is randomly drawn and determines both the facula lifetime and maximum radius.

\begin{figure}
    \centering
    \gridline{\includegraphics[width=0.5\textwidth]{figures/facula_model.png}}
    \gridline{\includegraphics[width=0.5\textwidth]{figures/facula_depth.pdf}}
    \script{facula_depth.py}
    \caption{
        ``Hot Wall'' model of faculae. Faculae structure causes their contrast to be dependent on their distance from the center of the disk. {\bf Top}: Depiction of a facula on the limb of a star. The hot wall is exposed to the observer causing the pore to appear bright. At disk center, the cool floor is most visible. {\bf Bottom}: The effects of depression depth and viewing angle on facula brightness. The 3D structure of faculae is most apparent when radius $\sim$ depth. A toy flux model was used to demonstrate the shape of these curves, but in practice their magnitudes depend on stellar spectral models.
        }
    \label{fig:fac_struct}
\end{figure}

\subsection{Flares \label{subsec:flares}}
Flares are an important source of stellar variability on short timescales. \texttt{vspec-vsm}'s flare lightcurve model is based on the \texttt{xoflares} package \citep{barclay2020}, which itself is based on empirically derived lightcurve shapes from \citet{davenport2016}. However, \texttt{xoflares} is designed to fit lightcurves in a single spectral band, so we add additional parameters to extend it to a multiwavelength lightcurve. We completely describe a flare using its temperature, total energy, full width at half maximum (FWHM), and the time of its peak. We model a flare as a hot, optically thin region above the photosphere that produces a blackbody spectrum. In this model, the temperature is constant, and the sharp rise and fall seen in the lightcurve is caused by the region's rapidly changing area. This simplification allows the total flare energy $E$ to be used as a normalization factor:
\begin{equation}
    \int_{-\infty}^{\infty}A\,dt = \frac{E}{\sigma T^4}
\end{equation}
where $A$ is the time-dependent area of the flare region, $T$ is the constant flare temperature, and $\sigma$ is the Stefan-Boltzmann constant. This allows flare lightcurves to be produced with a fixed bolometric luminosity. Because the temperature is fixed, the relevant quantity for modeling the flux of a flare in a given integration is the integrated time-area $\int A\,dt$. These quantities are computed numerically and fed to \vspec{} along with $T$ to produce spectra with the appropriate absolute flux.

Before simulating an observation, \vspec{} asks \texttt{vspec-vsm} to pre-compute a population of flares based on a power-law frequency-energy relationship similar to the results of \TJ{what data?} \citep{gao2022}:
\begin{equation} \label{eq:flare_freq}
    \log{(f/{\rm [day]})} = \beta + \alpha~\log{(E/{\rm [erg]})}
\end{equation}
where $f$ is the frequency of flares with energies $\ge E$. We compute the number of expected flares over a time duration and determine the number to create $N$ by a random Poisson draw. We then generate $N$ flares with energies determined by the quantile function:
\begin{equation}
    E = E_{\text{min}} (1-X)^{1/\alpha}
\end{equation}
where $E_{\text{min}}$ is the minimum considered flare energy and $X$ is a random number on the interval $[0,1)$. We set the default values $\alpha=-0.829$, $\beta=26.87$ from \citet{gao2022}, but these can be adjusted by the user. Figure \ref{fig:flare_freq} shows our simulation results compared to their power-law values.

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figures/flare_freq.pdf}
    \caption{
        {\bf Top}: Generated flare frequencies compared to those expected from \citet{gao2022}, generated using a 10,000 day simulation. The $y$-axis shows the frequency of flares with energies greater than or equal to the value of the $x$-axis. Error bars based on the square root of the number of observed flares as this is a Poisson process. {\bf Bottom}: Lightcurve of a star flaring with the same power-law slope as the top panel, but the intercept ($\beta$) has been increased by 0.3 for visual effect. The mean flare temperature is $9000$ K and the mean FWHM is $3$ hours.
        }
    \script{flare_freq.py}
    \label{fig:flare_freq}
\end{figure}

\subsection{Granulation}
Granulation is a source of stellar variability that arises from convection near the surface of the star. The result is a stellar surface that is not constant in temperature and that changes on very short timescales. Hydrodynamic simulations of stellar atmospheres \citep[e.g.][]{magic2014} describe hot granules of rising gas surrounded by cooler, sinking regions with a temperature a few percent lower than the nominal value.

We model granulation as a global process, and its effects are computed after the effects of spots and faculae. Of the remaining ``quiet'' photosphere (i.e. the regions not covered in spots of faculae), a fraction is computed to be part of the cool
inter-granule surface; the surface coverage of the cool region at any given time is computed by a Gaussian process (GP) using the {\sc tinygp} package \citep{foreman-mackey2024} following the methodology of \citet{gordon2020}. The GP uses a custom kernel function based on a power spectrum \citep{anderson1990,kallinger2014} to produce random changes in the granulation coverage.

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figures/granulation_lightcurve.pdf}
    \caption{
        Effect of granulation on the lightcurve of a star. This simulation assumes a mean
        10\% coverage by an inter-granule region whose temperature is 300K lower than the surrounding photosphere. It also assumes that the coverage value varies with a magnitude of 1\% on a timescale of 6 hours \TJ{find some constraints in Gordon paper}.
        }
    \script{granule_lc.py}
    \label{fig:gran_lc}
\end{figure}

\subsection{Transits}
\TJ{Change section to talk about limb darkening and other geometric considerations}
In the case of a transiting geometry, \texttt{vspec-vsm} is responsible for computing the properties of the portion of the stellar disk that is occulted. It does this by first projecting both the transiting planet and the visible part of the surface orthographically as they would be viewed by a distant observer. It then selects all the points that are near enough to the transit to be occulted. The code then iterates through each of those points and computes the fraction of it covered by the occulting circle using a 2D numerical integral. \TJ{Add a figure to show limb darkening lightcurve. Do a surface map like Fig 1 with a transit, then zoom in to show pixel coverage, then a lightcurve}

@@

\section{Simplified Climate Model}
\vspec{} allows the user to upload any desired GCM. However, for questions that focus on detectability,
the specific physics of the GCM are less important than properties like day/night temperature, surface pressure, or
the major absorbers in the atmosphere. To simplify these studies, \vspec{} has a built-in GCM that takes in just a
few parameters and produces a simple-yet-robust exoplanet climate. Simplicity allows the user to know everything about
the model and reduce surprises, but its basis in thermodynamics and robust energy balance make this model fit for publication-level science \TJ{is this really true? probably not}.
